### 2022-08-11 main/chapter_2ljets.tex
proofread results and conclusion; add disco region refs

With such a complicated experiment as \atlas, in a universe with such varied
particles as ours, there are many and varied factors of uncertainty to consider.
As described in Section~\ref{sec:searches_practice}, the effect of each
factor is studied to assign regularized variations to the Poisson expectations
of the histogram bins with a HistFactory model~\cite{cranmer2012histfactory}.
For the $\twoljets$ search, we use \histfitter\ to build and analyse those
models~\cite{Besjes_2015,baak2015histfitter}.

@@

These three alternative estimates are labelled `MC', `MM', `Mean', and
reported with their combination the final `Max' estimate in
Table~\ref{tab:2ljets_fnp_estimates}.
Each is described here.
\begin{itemize}
\item The `MC' estimate takes the simulated estimate that was previously reported
in Table~\ref{tab:2ljets_fnp_negative_mm}.
\item The `MM' estimate modifies the Matrix Method result,
also from Table~\ref{tab:2ljets_fnp_negative_mm}, by shifting its central
value to a small positive value chosen baselessly as $0.01$ and by adding
its statistical and systematic uncertainties linearly
(not in quadrature to conservatively cover a correlated variation)
to assign a high variation.
\end{itemize}
After I had proposed a combination of these two estimates, and long after
unblinding, internal reviewers suggested additionally using the statistical
properties of samples in looser selections.
This became the `Mean' estimate.
\begin{itemize}
\item The `Mean' follows an idea that the smallest assignable value should
be comparable to typical Matrix Method weights, and those typical values
are extracted from a region requiring pre-selection,
from Table~\ref{tab:2ljets_presel},
$\mll \in 71\textrm{--}101$,
$\mjj \in 60\textrm{--}100$,
$\njet \geq 2$, and
$\nbtag \leq 1$.
Its central value takes the mean weight of Matrix Method samples in this
selection, which is $\mu^\textrm{M} = 0.0202$.
Statistical errors are set to the root of the mean squared weight,
which is $\sigma^\textrm{M}_\textrm{stat} = 0.1132$,
and systematic errors are set as the larger absolute deviation on the mean
from the up and down variations in the Matrix Method weights, which comes out
$\sigma^\textrm{M}_\textrm{sys} = 0.0226$.
Finally, the `Mean' high variation is their linear sum;
$\mu^\textrm{M} + \sigma^\textrm{M}_\textrm{stat} +
\sigma^\textrm{M}_\textrm{sys} = 0.1560$.
\end{itemize}
All low variations are set to zero for these replacement FNP estimates, to
encode that the yield is compatible with zero, but more importantly to give
a slow extrapolation on the low side (from the small central value to zero)
and reduce the risk from non-physical negative yields in the HistFactory model.

@@

For their implementation in the background model in HistFactory, these
variations reduce to the relative scale factors that are presented in
Table~\ref{tab:2ljets_zjets_alt}.
Note that up and down variations of the sample comparison parameter do not
necessarily mean up and down variations of bin yields, but to the coupled
variations in directions given by the signs of the relative variations given in
the various regions.
These variations are evaluated according to a standard recipe prescribed by
the ATLAS SUSY group~\cite{atlas_twiki_susytheoretical} that says to
\begin{itemize}
\item evaluate the difference between the two samples $\delta_{12}$.
\item evaluate the statistical uncertainty on the alternative sample $\sigma_2$.
\item combine these components in quadrature, so
$\sigma_{12} = \sqrt{\delta_{12}^2 + \sigma_2^2}$, and
\item if the deviation is statistically significant, then preserve the direction
of the uncertainty, so $\sigma = \mathrm{sgn}(\delta_{12})\sigma_{12}$.
\end{itemize}
An `if A then B' statement does not define what to do if A is false, so
this recipe does not define what sign to give $\sigma$ if the deviation is not
statistically significant.
Since all variations have direction
--- they are either positive or negative or zero ---
we are unable to give no direction to any deviation, and we choose to
use the same, signed, definition of $\sigma$ for all regions.

@@

With this decision, post-fit results (such as the summary plot of
Figure~\ref{fig:2ljets_summary})
indeed show no significant deviation from data in SR-Low-a or SR-Low-b,
and this $\zjets$ sample comparison is the main reason why.
Internal reviewers were, there, confused as to why we mentioned a deficit in
SR-Low that did not appear in the results plots or tables.
Much to my ire, the paper editors therefore removed all mentions of an SR-Low
deficit from the paper.
Despite bring real and significant, this deficit was erased from public view,
and to my knowledge, is only reported through inner details of
the HistFactory models shared in our \hepdata\ archive~\cite{hepdata.116034}
(and now, I hope, this thesis).

@@

% pro
Naturally, \atlas\ has an unblinding procedure that requires committee
review and approval based on internal documentation.
In the \atlas\ SUSY world, searches are nominally performed blind to signal
regions.
Control and validation regions are consumed in the design and modelling of
signal regions, so need to be inspected.
Control and validation region data are inspected, but this is not a problem
since they should also have negligibly small yields (contaminations) from
interesting signal models.
With negligible signal contaminations, control and validation region data can
be consumed to constrain the background modelling in addition to all other that
on which the Standard Model stands, and without being influenced by the signal
models to be tested in signal regions.

The \atlas\ SUSY approval procedure~\cite{atlas_twiki_susyapprove}
requires various criteria including, and not limited to, that the search has:
\begin{itemize}
\item defined all control, validation and signal regions,
\item evaluated signal contaminations in control and validation regions,
\item estimated all regions' simulated backgrounds,
\item outlined procedures to estimate non-simulated (data-driven) backgrounds,
\item evaluated variations due to experimental uncertainties in all regions
according to recommended procedures,
\item stated clearly how all other (theoretical) uncertainties will be treated.
\end{itemize}
So many aspects must be finalized blind, and for those that are not
finalized blind must have procedures defined such that they are not
influenced by the signal region data.
Encouragingly, disallowing dependence of procedures on signal region data
eliminates one major source of unblinding error,
and requiring knowledge of signal contaminations supports the use of control
and validation region data.

% con
Unfortunately, these requirements limit only \emph{when} data-driven estimation
procedures are defined, and do not limit the \emph{what} those procedures do.
Indeed, the rules explicitly allow for methods like the Matrix Method
(described in Section~\ref{sec:2ljets_matrix_method})
that use signal region data to affect the background `estimates' in signal
regions in violation of blinding principles.
No effective research would employ an estimation procedure that clearly set
a prediction directly from the data it is supposed to predict, but our use of
the Matrix Method is proof by example that sufficiently obfuscated violations
are allowed here.

A second unfortunate flaw is that these rules were not strictly enforced.
Although we performed the approval procedure as written, some of its promises
were later violated, at the request of senior committees, without consequence.
The data-driven FNP estimation procedure should have been frozen, but was in
fact extended with the `Mean' method after unblinding,
as seen in Section~\ref{sec:2ljets_matrix_method_ewk}).
Theoretical uncertainties were always to be treated according to the
\atlas\ SUSY recommendations~\cite{atlas_twiki_susytheoretical}, but those
recommendations have ambiguities that were resolved after unblinding,
notably for the alternative $\zjets$ sample comparison as discussed in
Section~\ref{sec:2ljets_zjets_alt}, among other less impactful sample
comparisons.

@@

When first introduced to this search project, I was shown plots resembling
(but not identical to) those in Figure~\ref{fig:2ljets_prepre}, which present
data and pre-fit backgrounds in loose selections.
Such plots were discussed in group meetings and used to catch bugs in the
background simulations.
But they unblind the data, notably in high-$\met$ where we expect signals.
When commenting on this, I was told that it does not count as unblinding
because we are in a loose selection.
Any signal region must be within this loose selection, and those
highest bins contain only a few events that agree with the backgrounds well.
From such plots, we knew before designing the regions that either there was no
excess at high $\met$ or that a signal region excess must coincide with a
deficit in its surrounding phase space (or vice versa).
Since we fill surrounding phase space with validation regions, one could in
fact have predicted an SR-High deficit from the subtle excess in VR-High
that is shown in Figure~\ref{fig:2ljets_high_mjj_vrhigh}.%
\footnote{%
Lacking this intelligence, at the time of unblinding I was actually concerned
that we were missing a background (perhaps from pileup jets) that would
explain low-$\mjj$ excess in VR-High, and might lead to a false excess
in SR-High.
My naive past self therefore found the observed zero events in SR-High-8
quite hilarious.
}
If blinding is to be taken seriously, then restrictions must be placed on the
use of such pre-selection plots.

% unblinded results
\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figures/2ljets_summary_unblinding_log.png}
\caption[
First unblinded summary table from 28th September 2020
]{%
First unblinded summary table from 28th September 2020.
Backgrounds are fitted to data in control and signal regions.
Errors include all variations that were evaluated at the time.
Data and backgrounds were later reproduced with updated software and tighter
lepton definitions, leading to a general reduction in yields.
}
\label{fig:2ljets_summary_unblinding}
\end{figure}

Our initial unblinded results are displayed in the summary plot of
Figure~\ref{fig:2ljets_summary_unblinding}.
This figure includes all backgrounds and variations that had been evaluated
at the time, and is included for comparison against the final result
that has been displayed in Figure~\ref{fig:2ljets_summary}.
We knew when unblinding that our simulations and data needed to be reprocessed
with the lepton definitions given in Section~\ref{sec:2ljets_events_leptons},
which had previously required harder baseline leptons,
and had expected only minor changes.
As seen from these two figures, the reprocessing reduced our yields of data
slightly; most noticeably, the one event in SR-$\llbb$ disappeared.
Other than aesthetic differences, post-unblinding activities have had little
impact.
The reported significance of deficits in SR-Low-a and SR-Low-b has reduced,
and the shrinking $\zjets$ background has caused a $1$-sigma excess in VR-Low.
Excesses have also grown in VR-High-* regions.
Little has changed in the SR-High-8 bins with zero data since, as explained in
Section~\ref{sec:2ljets_diboson_alt}, the added \diboson\ sample comparison
is ineffective.

Despite my concerns expressed in this section, these post-unblinding changes
are relatively minor.
They do not change the broad message that there are no excesses here, and at
most amount to a misleadingly small background estimate in SR-Low bins if the
post-fit yields that we present are misinterpreted as predictions.
Pre-fit yields, which can plausibly be interpreted as predictions, are
presented with other extra results in the next section.

@@

\paragraph{Summary}
As I have argued incessantly, pre-fit predictions are useful because they can
actually be predictions to be tested by the data.
Our post-fit summary plot has previously been displayed in
Figure~\ref{fig:2ljets_summary}.
Its alternative version that does not fit to signal region data is shown here
in Figure~\ref{fig:2ljets_summary_no_sr}.
Since the significance measure is meaningful for this pre-signal-region plot,
we see that despite the inflated uncertainties in SR-Low, its bins still have
substantial deficits.
Validation region data are notably better predicted by this prediction;
as the post-fit result tunes itself to signal region data, coupled variations
lead to worse predictions in the validation regions (that are not included
directly in either fit).

@@

\paragraph{Correlations}
All parameters are initially independent.
If the \heplikelihood\ is related to a probability distribution of parameters,
however, then they gain interdependencies when coupled through the likelihoods
of control and signal region data.
The correlation matrix of parameters, estimated from an approximate Hessian
matrix at the best fit \heplikelihood, is displayed in
Figure~\ref{fig:2ljets_fit_corrmatrix}.
To reduce this matrix to a more readable size, only the most correlated
parameters are shown.

Mixed contributions of $\ttbar$, $\zjets$ and \diboson\ in the control regions
leads to anticorrelations between their normalization factors
``\texttt{mu\_*}'' that are seen in the bottom right corner of this correlation
plot.
Since statistical errors ``\texttt{gamma\_stat\_*}'' in control regions are not
absorbed into the sample-specific normalization factors, they are also
anticorrelated with normalization factors.
Similarly, the extra uncertainty for $\zjets$ in CR-tt causes an
anticorrelation between its parameter ``\texttt{alpha\_Zjets\_bextrap}''
and the $\ttbar$ normalization.
The failure of statistical precision in SR-Low-2 causes large systematic
variations that appear as the anticorrelations between
\texttt{gamma\_stat\_SRLow2\_cuts\_bin\_0} and numerous experimental
uncertainties.
Correlations between the $\zyjets$ normalization and $\ttbar$ uncertainties
are due to large variations in the alternative $\ttbar$ samples in \crdy.
Electron identification variations ``\texttt{alpha\_el\_chargeid\_*}'' have
some impact in SR-Low,  so are correlated with the $\zjets$ alternative sample
comparison ``\texttt{alpha\_Zjets\_alt}''.
Note that the up and down variations of parameters need not correspond to up
and down variations of yields in any region, so the signs of these correlations
are no necessarily meaningful.

@@

Parameters named ``\texttt{$\alpha$\_*}'' are centred on zero with unit initial width,
and others are initially relative to $1$ with variable error bars.
Internally, plots like this are considered good if all parameters sit on
the central vertical line, with the motivations that normalization factors
should be close to $1$ (the predicted cross-sections are accurate) and that
our data should not compete with the precision of the studies that defined
our systematic variations.
That is, if we constrain a parameter by squeezing its error bar, then our data
could be better used to calibrate that parameter throughout the Collaboration.

I added the red arrow to indicate the $\zjets$ alternative sample comparison,
named ``\texttt{$\alpha$\_Zjets\_alt}'' here, that has disappeared off the
right-hand edge due to its best-fit value of $1.86 \pm 0.69$.
(see Section~\ref{sec:2ljets_zjets_alt}).
Its anticorrelation with the electron charge identification variations is
explained by their movement to negative values;
positive \texttt{$\alpha$\_Zjets\_alt} and negative
\texttt{$\alpha$\_el\_chargeid\_*} both reduce yields in SR-Low.

@@

\paragraph{Contours}
Exclusion contours, which are displayed in
Figures~\ref{fig:2ljets_contours_c1n2} and~\ref{fig:2ljets_contours_gmsb},
derive from fits with models that extend the background-only model with
additive signal samples.
To reduce the computational cost of building the hundreds of models%
\footnote{%
This is worded precisely. The dominant cost was not in fitting the models, nor
in histogramming the data (thanks to histogram caching), but in forming the
HistFactory model from those histograms.
I do not know why that model-building step took days of computer time per
model, but it did reduce to only hours after upgrading the \histfitter\ version
from \textsc{v0.64.0} to \textsc{v0.66.0}.
}%
across these grids, we prune the number of small irrelevant parameters
with \histfitter's pruning~\cite{histfittergithub} mechanism in ``method 2''.
This pruning method removes variations if the difference between their up and
down values do not exceed a threshold.
With pruning enabled, its default threshold is set to $1\%$ (relative to the
central yield), but to reduce its impact we use a threshold of $0.1\%$.
Validation studies comparing each of these pruning thresholds, against the
un-pruned version, and against a custom pruning implemented by myself, found no
significant differences.%
\footnote{%
Pruning is, however, the reason our signal patches (that describe differences
from background-only to signal models) in our serialized \hepdata\ models are
so large; they implement many small changes due to different decisions made by
the pruning algorithm.
}

@@

To some extent, our published exclusion contours are an artistic choice.
What we know are the values of statistics in an irregular grid of
signal points,
and all contours are painted by heuristic interpolation algorithms.
We follow standard \atlas\ SUSY conventions and perform that interpolation
with the \textsc{Scipy} implementation of
Radial Basis Functions~\cite{2020SciPy-NMeth,fasshauer2007meshfree},
interpolating significance-transformed $\cls$ values.
That is, we interpolate in $\cls$ values transformed through the inverse
cumulative distribution of the standard normal distribution.
Contours are drawn where this interpolated function meets
$\cls = 0.05$.
\emph{%
Thanks to Larry~Lee for sharing scripts that were adapted to prepare the
contours in our results.%
}

This interpolation algorithm has some tunable parameters, for which all
contours of the $\twoljets$-electroweak search use
\texttt{epsilon=10},
\texttt{smoothing=1},
and the ``multiquadric'' radial basis function.
These settings were tuned to keep the thumbs-up wobble in the C1N2 off-shell
region of Figure~\ref{fig:2ljets_contours_c1n2},
which is physical and due to our increased sensitivity for mass splittings
below $m(W)$,
while sufficiently smoothing wobbles elsewhere.
Some unattractive wobbles remain at high GMSB masses in
Figure~\ref{fig:2ljets_contours_gmsb}, but I decided that it was preferable to
keep the simplicity of having the same interpolation settings for both
contours.

Our initial contours showed more glaring and unphysical wobbles.
Those erroneous wobbles were removed by repeating the fits of anomalous signal
points until they behaved.
In total, this hack was used for two points near the highest
$m(\chargino_1,\neutralino_2)$ edge in C1N2, and four across the whole
GMSB spectrum.
Different random seeds causing failed fits explain this unstable behaviour,
and I failed to control those random seeds when running the fits through
\histfitter.

Auxiliary results published in our \hepdata\ archive~\cite{hepdata.116034}
include the observed upper limits on the signal cross-section at each point.
We additionally display the $\cls$ values, from which the observed contour
derives, in Figure~\ref{fig:2ljets_fit_contours_cls}.
By zooming in, one can see that not all points with $\cls < 0.05$ are
labelled as excluded;
this is direct a result of our smoothing choices.

@@

@@

Many test statistic distributions are involved here --- two (background-only
and signal-plus-background) at each scanned point, and one additional
distribution per region (with yet another test statistic) for the discovery
$p$-value $p_{s\!=\!0}$.
From these, a few examples are presented in
Figure~\ref{fig:2ljets_test_stat_distributions}
to investigate curious features of the results.

>>>>

With such a complicated experiment as \atlas, in a universe with such varied
particles as ours, there are many and varied factors of uncertainty to consider.
As described in Section~\ref{sec:searches_practice}, the effect of each
factor is studied to assign regularized variations to the Poisson expectations
of the histogram bins with a \histfactory\ model~\cite{cranmer2012histfactory}.
For the $\twoljets$ search, we use \histfitter\ to build and analyse those
models~\cite{Besjes_2015,baak2015histfitter}.

@@

These three alternative estimates are labelled `MC', `MM', `Mean', and
reported with their combination the final `Max' estimate in
Table~\ref{tab:2ljets_fnp_estimates}.
Each is described here.
\begin{itemize}
\item The `MC' estimate takes the simulated estimate that was previously reported
in Table~\ref{tab:2ljets_fnp_negative_mm}.
\item The `MM' estimate modifies the Matrix Method result,
also from Table~\ref{tab:2ljets_fnp_negative_mm}, by shifting its central
value to a small positive value chosen baselessly as $0.01$ and by adding
its statistical and systematic uncertainties linearly
(not in quadrature to conservatively cover a correlated variation)
to assign a high variation.
\end{itemize}
After I had proposed a combination of these two estimates, and long after
unblinding, internal reviewers suggested additionally using the statistical
properties of samples in looser selections.
This became the `Mean' estimate.
\begin{itemize}
\item The `Mean' follows an idea that the smallest assignable value should
be comparable to typical Matrix Method weights, and those typical values
are extracted from a region requiring pre-selection,
from Table~\ref{tab:2ljets_presel},
$\mll \in 71\textrm{--}101$,
$\mjj \in 60\textrm{--}100$,
$\njet \geq 2$, and
$\nbtag \leq 1$.
Its central value takes the mean weight of Matrix Method samples in this
selection, which is $\mu^\textrm{M} = 0.0202$.
Statistical errors are set to the root of the mean squared weight,
which is $\sigma^\textrm{M}_\textrm{stat} = 0.1132$,
and systematic errors are set as the larger absolute deviation on the mean
from the up and down variations in the Matrix Method weights, which comes out
$\sigma^\textrm{M}_\textrm{sys} = 0.0226$.
Finally, the `Mean' high variation is their linear sum;
$\mu^\textrm{M} + \sigma^\textrm{M}_\textrm{stat} +
\sigma^\textrm{M}_\textrm{sys} = 0.1560$.
\end{itemize}
All low variations are set to zero for these replacement FNP estimates, to
encode that the yield is compatible with zero, but more importantly to give
a slow extrapolation on the low side (from the small central value to zero)
and reduce the risk from non-physical negative yields in the \histfactory\
model.

@@

For their implementation in the background model in \histfactory, these
variations reduce to the relative scale factors that are presented in
Table~\ref{tab:2ljets_zjets_alt}.
Note that up and down variations of the sample comparison parameter do not
necessarily mean up and down variations of bin yields, but to the coupled
variations in directions given by the signs of the relative variations given in
the various regions.
These variations are evaluated according to a standard recipe prescribed by
the ATLAS SUSY group~\cite{atlas_twiki_susytheoretical} that says to
\begin{itemize}
\item evaluate the difference between the two samples $\delta_{12}$.
\item evaluate the statistical uncertainty on the alternative sample $\sigma_2$.
\item combine these components in quadrature, so
$\sigma_{12} = \sqrt{\delta_{12}^2 + \sigma_2^2}$, and
\item if the deviation is statistically significant, then preserve the direction
of the uncertainty, so $\sigma = \mathrm{sgn}(\delta_{12})\sigma_{12}$.
\end{itemize}
An `if A then B' statement does not define what to do if A is false, so
this recipe does not define what sign to give $\sigma$ if the deviation is not
statistically significant.
Since all variations have direction
--- they are either positive or negative or zero ---
we are unable to give no direction to any deviation, and we choose to
use the same, signed, definition of $\sigma$ for all regions.

@@

With this decision, post-fit results (such as the summary plot of
Figure~\ref{fig:2ljets_summary})
indeed show no significant deviation from data in SR-Low-a or SR-Low-b,
and this $\zjets$ sample comparison is the main reason why.
Internal reviewers were, there, confused as to why we mentioned a deficit in
SR-Low that did not appear in the results plots or tables.
Much to my ire, the paper editors therefore removed all mentions of an SR-Low
deficit from the paper.
Despite bring real and significant, this deficit was erased from public view,
and to my knowledge, is only reported through inner details of
the \histfactory\ models shared in our \hepdata\ archive~\cite{hepdata.116034}
(and now, I hope, this thesis).

@@

% pro
Naturally, \atlas\ has an unblinding procedure that requires committee
review and approval in accordance with internal documentation.
In the \atlas\ SUSY world, searches are nominally performed blind to signal
regions.
For these searches, their control and validation regions must consumed in the
design and modelling of signal regions, so they must be inspected.
This is not a problem since control and validation regions should also
have negligibly small yields (contaminations) from interesting signal models,
With negligible signal contaminations, control and validation region data can
be consumed to constrain the background modelling, in composition with to all
other data on which the Standard Model stands, and without being influenced by
possible data from the target signal models that are to be tested in signal
regions.

Before allowing unblinding, the \atlas\ SUSY approval
procedure~\cite{atlas_twiki_susyapprove} requires various criteria including,
and not limited to, that a search has:
\begin{itemize}
\item defined all control, validation and signal regions,
\item evaluated signal contaminations in control and validation regions,
\item estimated all regions' simulated backgrounds,
\item outlined procedures to estimate non-simulated (data-driven) backgrounds,
\item evaluated variations due to experimental uncertainties in all regions
according to recommended procedures,
\item stated clearly how all other (theoretical) uncertainties will be treated.
\end{itemize}
Many aspects of a search must therefore be finalized blind, and those aspects
that are not finalized blind must have procedures defined such that they are
not influenced by the signal region data.
Encouragingly, this disallowance of dependence of procedures on signal region
data eliminates one major source of unblinding error,
and requiring knowledge of signal contaminations supports the use of control
and validation region data.

Unfortunately, these requirements limit only \emph{when} data-driven estimation
procedures are defined, and do not limit the \emph{what} those procedures
\emph{do}.
The rules explicitly allow for methods like the Matrix Method
(described in Section~\ref{sec:2ljets_matrix_method})
that use signal region data to determine their `estimates' in signal
regions, in violation of basic blinding principles.
Only the procedure is blind to data, not its action.
No scientific research would employ an estimation procedure that clearly set
its prediction directly from the data it is supposed to predict, but our use of
the Matrix Method is proof by example that sufficiently obfuscated violations
are permitted here.

A second unfortunate flaw is that no mechanism enforces these rules after
unblinding.
Although we performed the approval procedure as written, some of its promises
were later violated on the instruction of committee decisions and without
negative consequences.
For example, by these rules the data-driven FNP estimation procedure should
have been fixed, but it in fact extended with the `Mean' method after
unblinding, as described in Section~\ref{sec:2ljets_matrix_method_ewk}.
Theoretical uncertainties were always to be treated according to the
\atlas\ SUSY recommendations~\cite{atlas_twiki_susytheoretical}, but those
recommendations have ambiguities that were resolved after unblinding.
For example, the decision to include the various alternative sample comparisons
was made after unblinding, and debates on our treatment of the alternative
$\zjets$ sample comparison made direct reference to the anomalous unblinded
signal region data, as described in Section~\ref{sec:2ljets_zjets_alt}.
On balance, however, we are fortunate that the many post-unblinding validation
studies found no actionable flaws. We cannot know what would have changed if a
validation study had found reasons to further adapt our modelling to data after
unblinding.

@@

When first introduced to this search project, I was shown plots resembling
(but not identical to) those in Figure~\ref{fig:2ljets_prepre}, which present
data and pre-fit backgrounds in loose selections.
Such plots were discussed in group meetings and used to catch bugs in the
background simulations.
But they unblind the data, notably in high-$\met$ where we expect signals.
When commenting on this, I was told that it does not count as unblinding
because we are in a loose selection.
However, any signal region must be within this loose selection, and those
highest bins contain only a few events that agree with the backgrounds well.
From such plots, we therefore knew before designing the regions that either
there was no excess at high $\met$ or that a signal region excess must
coincide with a deficit in its surrounding phase space (or vice versa).
Since we fill surrounding phase space with validation regions, an excess in a
validation region therefore predicts a deficit in its signal region.
One could have predicted an SR-High deficit from the excess in VR-High
that is shown in Figure~\ref{fig:2ljets_high_mjj_vrhigh}.%
\footnote{%
Lacking this intelligence, at the time of unblinding I was actually concerned
that we were missing a background (perhaps from pileup jets) that would
explain low-$\mjj$ excess in VR-High, and might lead to a false excess
in SR-High.
My naive past self therefore found the observed zero events in SR-High-8
quite hilarious.
}
For blinding to be done properly, restrictions must be placed on the usage of
such pre-selection plots.

% unblinded results
\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figures/2ljets_summary_unblinding_log.png}
\caption[
First unblinded summary table from 28th September 2020
]{%
First unblinded summary table from 28th September 2020.
Backgrounds are fitted to data in control and signal regions.
Errors include all variations that were evaluated at the time.
Data and backgrounds were later rederived with updated software and tighter
lepton definitions, leading to a general reduction in yields.
}
\label{fig:2ljets_summary_unblinding}
\end{figure}

Our initial unblinded results are displayed in the summary plot of
Figure~\ref{fig:2ljets_summary_unblinding}.
This figure includes all backgrounds and variations that had been evaluated
at the time, and is included for comparison against the final result
that has been displayed in Figure~\ref{fig:2ljets_summary}.
We knew when unblinding that our simulations and data needed to be reprocessed
with the lepton definitions given in Section~\ref{sec:2ljets_events_leptons}.
Our previous baseline lepton definition had required a larger $\pt$,
and we had expected this change to make minimal differences.
As seen from these two figures, the reprocessing reduced our yields of data
slightly; perhaps most noticeably, the one event in SR-$\llbb$ disappeared.
Other than aesthetic differences, and post-unblinding activities have on
balance had relatively little impact to the post-fit results.
The reported significance of deficits in SR-Low-a and SR-Low-b has reduced,
and the shrinking $\zjets$ background has caused a $1$-sigma excess in VR-Low.
Excesses have also grown in VR-High-* regions as the model has further
adapted towards the signal region data.
Little has changed in the SR-High-8 bins with zero data since, as explained in
Section~\ref{sec:2ljets_diboson_alt}, the extra \diboson\ sample comparison
is impotent.

Despite my concerns expressed in this section, these post-unblinding changes
are relatively minor.
They do not change the broad message that there are no excesses here, and at
most amount to a misleadingly small background estimate in SR-Low bins if the
post-fit yields that we present are misinterpreted as predictions.
Pre-fit yields, which can plausibly be interpreted as more like predictions,
are presented with other extra results in the next section.

@@

\paragraph{Summary}
As I have argued quite incessantly throughout this thesis, pre-fit predictions
are useful because they can be predictions to be tested by the data.
Our post-fit summary plot has previously been displayed in
Figure~\ref{fig:2ljets_summary}, and an alternative version that does not fit
to signal region data is displayed in Figure~\ref{fig:2ljets_summary_no_sr}.
The significance measure is more meaningful for this pre-signal-region plot
since it expects pre-fit uncertainties, and we see here that despite the
inflated $\zjets$ uncertainties in SR-Low its bins still have substantial
deficits.
Validation region data are notably better predicted pre-fit ---
as the post-fit result tunes itself to signal region data, coupled variations
lead to worse predictions in the validation regions since their likelihoods are
not included directly in any fits.

@@

\paragraph{Correlations}
All parameters of our \histfactory\ models are initially independent.
If the \heplikelihood\ is related to a probability distribution of parameters,
however, then they gain interdependencies when coupled through the likelihoods
of control and signal region data.
The correlation matrix of parameters, estimated from an approximate Hessian
matrix at the best fit \heplikelihood, is displayed in
Figure~\ref{fig:2ljets_fit_corrmatrix}.
To reduce this matrix to a more readable size, only the most correlated
parameters are shown.

In this correlation matrix, mixed contributions of $\ttbar$, $\zjets$ and
\diboson\ in the control regions leads to anticorrelations between their
normalization factors
``\texttt{mu\_*}'' that are seen in the bottom right corner of this correlation
plot.
Since statistical errors ``\texttt{gamma\_stat\_*}'' in control regions are not
absorbed into the sample-specific normalization factors, they are also
anticorrelated with normalization factors.
Similarly, the extra uncertainty for $\zjets$ in CR-tt causes an
anticorrelation between its parameter ``\texttt{alpha\_Zjets\_bextrap}''
and the $\ttbar$ normalization.
The failure of statistical precision in SR-Low-2 causes large systematic
variations that appear as the anticorrelations between its statistical error
\texttt{gamma\_stat\_SRLow2\_cuts\_bin\_0} and numerous experimental
uncertainties.
Correlations between the $\zyjets$ normalization and $\ttbar$ uncertainties
are due to large variations in the alternative $\ttbar$ samples in \crdy.
Electron identification variations ``\texttt{alpha\_el\_chargeid\_*}'' have
some impact in SR-Low,  so are correlated with the $\zjets$ alternative sample
comparison ``\texttt{alpha\_Zjets\_alt}''.

Note that the up and down variations of parameters need not correspond to up
and down variations of yields in any region, so the signs of these
(anti)correlations are no necessarily meaningful.

@@

Parameters named ``\texttt{$\alpha$\_*}'' are centred on zero with unit initial width,
and others are initially relative to $1$ with variable error bars.
Within the \atlas\ SUSY group, plots like this are considered good if all
parameters sit on the central vertical line, with the motivations that
normalization factors should be close to $1$ (the predicted cross-sections are
accurate) and that our data should not compete with the precision of the
studies that defined our systematic variations.
That is, if we constrain a parameter by squeezing its error bar, then our data
could be better used to calibrate that parameter throughout the Collaboration.

Here, I added the big red arrow to indicate the $\zjets$ alternative sample
comparison, named ``\texttt{$\alpha$\_Zjets\_alt}'', that has disappeared off
the right-hand edge due to its best-fit value of $1.86 \pm 0.69$.
(see Section~\ref{sec:2ljets_zjets_alt}).
Its anticorrelation with the electron charge identification variations is
explained by their movement to negative values:
positive \texttt{$\alpha$\_Zjets\_alt} and negative
\texttt{$\alpha$\_el\_chargeid\_*} both reduce yields in SR-Low.

@@

\paragraph{Contours}
Exclusion contours, displayed in
Figures~\ref{fig:2ljets_contours_c1n2} and~\ref{fig:2ljets_contours_gmsb},
derive from fits with models that extend the background-only model with
additive signal samples.
To reduce the computational cost of building the hundreds of models%
\footnote{%
This is worded precisely. The dominant cost was not in fitting the models, nor
in histogramming the data (thanks to caching), but in forming the
\histfactory\ model from those histograms.
I do not know why that model-building step took days of computer time per
model, but it did reduce to only hours after upgrading the \histfitter\ version
from \textsc{v0.64.0} to \textsc{v0.66.0}.
}
across the signal parameter planes, we prune the number of small irrelevant
parameters with \histfitter's pruning~\cite{histfittergithub} mechanism in
``method 2''.
This pruning method removes variations if the difference between their up and
down values do not exceed a threshold.
With pruning enabled, its default threshold is set to $1\%$ (relative to the
central yield), but to minimize concerns about its impact we use a threshold
of $0.1\%$.
Validation studies that compared each of these pruning thresholds, the
un-pruned version, and a custom pruning implemented by myself, found no
significant differences to post-fit results.%
\footnote{%
Pruning is, however, the reason our signal patches (which encode differences
from background-only to signal models) in our serialized \hepdata\ models are
so large --- they implement many small changes due to different decisions made
by the pruning algorithm.
}

@@

To some extent, our published exclusion contours are an artistic choice.
What we know are the values of statistics in an irregular grid of
signal points,
and all contours are painted by heuristic interpolation algorithms.
We follow standard \atlas\ SUSY conventions and perform that interpolation
with the \textsc{Scipy} implementation of
Radial Basis Function~\cite{2020SciPy-NMeth,fasshauer2007meshfree}
interpolation, applied to significance-transformed $\cls$ values.
That is, we interpolate in $\cls$ values transformed through the inverse
cumulative distribution of the standard normal distribution.
Contours are drawn where this interpolated function meets the equivalent of
$\cls = 0.05$.
\emph{%
Thanks to Larry~Lee for sharing scripts that were adapted to prepare the
contours in our results.%
}

This interpolation algorithm has some tunable parameters, for which after some
experimentation I decided that all contours of the $\twoljets$-electroweak
search use
\texttt{epsilon=10},
\texttt{smoothing=1},
and the ``multiquadric'' radial basis function.
These settings were tuned to keep the `thumbs-up' wobble in the C1N2 off-shell
region of Figure~\ref{fig:2ljets_contours_c1n2},
which is physical and due to our increased sensitivity for some mass splittings
below $m(Z)$,
while sufficiently smoothing wobbles elsewhere.
Some unattractive wobbles remain at high GMSB masses in
Figure~\ref{fig:2ljets_contours_gmsb}, but I decided that it was preferable to
keep the simplicity of having the same interpolation settings for both
contours.

@@

Our initial contours showed more glaring and unphysical wobbles.
Those erroneous wobbles were removed by repeating the exclusion fits for signal
points with anomalous results until their results appeared to behave.
In total, this hack was used for two points near the highest
$m(\chargino_1,\neutralino_2)$ edge in C1N2, and four across the whole
GMSB spectrum.
Different random seeds in the optimization algorithms explain this unstable
behaviour; they clearly affect the fit results, and my attempts control all
random seeds failed when executing these fits through \histfitter.

Auxiliary results published in our \hepdata\ archive~\cite{hepdata.116034}
include the observed upper limits on the signal cross-section at each point.
We additionally display the $\cls$ values, from which the observed contour
derives, in Figure~\ref{fig:2ljets_fit_contours_cls}.
By zooming in, one can see that not all points with $\cls < 0.05$ are
labelled as excluded;
this is direct a result of our artistic interpolation choices.

@@

Many test statistic distributions are involved in these results ---
two (background-only and signal-plus-background) at each scanned point,
and one additional distribution per region for the discovery
$p$-value $p_{s\!=\!0}$.
Some examples of these distributions are displayed in
Figure~\ref{fig:2ljets_test_stat_distributions}
to help investigate curious features of the results.
